---
title: "Hands-on_Ex02"
editor: visual
---

## Overview

In this hands-on exercise, I learn how to plot functional and truthful choropleth maps by using r packages called **tmap** package.

Beside **tmap** package, four other R packages will be used. They are:

-   [**readr**](https://readr.tidyverse.org/) for importing delimited text file,

-   [**tidyr**](https://tidyr.tidyverse.org/) for tidying data,

-   [**dplyr**](https://dplyr.tidyverse.org/) for wrangling data and

-   [**sf**](https://cran.r-project.org/web/packages/sf/) for handling geospatial data.

```{r}
#pacman::p_load(sf,tmap, tidyverse)
library(sf)
library(tidyverse)
library("tmap")
```

## Import Geospatial Data into R

```{r}
mpsz <- st_read(dsn="data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

```{r}
mpsz
```

## Import attribute data into R

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

```{r}
popdata %>% head()
```

## Data preparation

Prepare a data table with year 2020 values. The data table should include the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

-   YOUNG: age group 0 to 4 until age groyup 20 to 24,

-   ECONOMY ACTIVE: age group 25-29 until age group 60-64,

-   AGED: age group 65 and above,

-   TOTAL: all age group, and

-   DEPENDENCY: the ratio between young and aged against economy active group

### 
Data Wrangling

-   *pivot_wider()* of **tidyr** package, and

-   *mutate()*, *filter()*, *group_by()* and *select()* of **dplyr** package

```{r}
popdata2020 <- popdata %>% 
  filter(Time == 2020) %>% 
  group_by(PA,SZ,AG) %>% 
  summarise(POP = sum(Pop)) %>% 
  ungroup() %>% 
  pivot_wider(names_from = AG,values_from = POP) %>% 
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>% 
  mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+rowSums(.[13:15])) %>%
  mutate(AGED = rowSums(.[16:21])) %>%
  mutate(TOTAL = rowSums(.[3:21])) %>% 
  mutate(DEPENDENCY = (YOUNG + AGED)/`ECONOMY ACTIVE`) %>% 
  select(PA,SZ,YOUNG,`ECONOMY ACTIVE`, AGED, TOTAL, DEPENDENCY)
  
  
popdata2020 
```

### Joining the attribute data and geospatial data

Convert PA and SZ fields in popdata2020 into all uppercase

```{r}
popdata2020 <- popdata2020 %>% 
  mutate_at(.var= vars(PA,SZ),
            .funs = funs(toupper)) %>%
  filter(`ECONOMY ACTIVE` >0)

```

Next, *left_join()* of **dplyr** is used to join the geographical data and attribute table using planning subzone name e.g. *SUBZONE_N* and *SZ* as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")

```

## Choropleth Mapping Geospatial Data Using tmap

Two approaches can be used to prepare thematic map using *tmap*, they are:

-   Plotting a thematic map quickly by using *qtm()*.

-   Plotting highly customisable thematic map by using tmap elements.

### Plotting a choropleth map quickly using qtm()

```{r}
top_n(mpsz_pop2020, -15,`DEPENDENCY`)

```

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020,
    fill = "DEPENDENCY")
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```
