---
title: "Hands-on_Ex04"
editor: visual
---

In this hands-on exercises, I will be

1\. computing Global (GSA) by using spdep package.

-   plot Moran Scatterplot

-   compute and plot spatial correlogram

2\. computing Local measure of Spatial Autocorrelation (LISA) to detect clusters and outliers by using spdep packages

3\. computing Getis-Ord's Gi-statistics for detecting hot spot or/ and cold spot area by using spdep package

4\. visualising the analysis output by using tmap package

## 4.2 Getting Started

### 4.2.1 The analytical question

-   is there equal distribution of development in the province?

-   if answer is NO, we ask "is there sign of clustering?" (GLOBAL spatial autocorrelation)

-   if YES, "Where are the clusters" (LOCAL spatial autocorrelation)

In this case study, we are interested to examine the spatial pattern of a selected development indicator (i.e. GDP per capita) of Hunan Provice, People Republic of China.

### 4.2.2 The Study Area and Data

Two data sets will be used in this hands-on exercise:

1.  geospatial data: Hunan province administrative boundary layer at county level in ESRI shapefile format
2.  aspatial data: Hunan_2012.csv containing local development indicators

### 4.2.3 Setting the Analytical Tools

Packages we need:

-   sf is use for importing and handling geospatial data in R,

-   tidyverse is mainly use for wrangling attribute data in R,

-   spdep will be used to compute spatial weights, global and local spatial autocorrelation statistics, and

-   tmap will be used to prepare cartographic quality chropleth map.

```{r}
library(sf)
library(spdep)
library(tmap)
library(tidyverse)
```

## 4.3 Getting the Data Into R Environment

### 4.3.1 Import shapefile into r environment

```{r}
hunan <-st_read(dsn='data/geospatial',
                layer= 'Hunan')
```

\*data is in WSG84 format, not projected yet. If we are going to extract the data, better not to st_transform() to project coordinate system first.

### 4.3.2 Import csv file into r environment

```{r}
hunan2012 <- read_csv('data/aspatial/Hunan_2012.csv')
```

### 4.3.3 Performing relational join

```{r}
hunan <- left_join(hunan, hunan2012)
```

```{r}
str(hunan)
```

```{r}
 hunan %>% 
  select(1:4, 7, 15)
```

### 4.3.4 Visualising Regional Development Indicator

Now we will prepare basemap and a choropleth map showing the distribution of GDPPC 2012 by using qtm() of tmap package

\*tm_fill() 's n refer to the number of equal intervals

```{r}
equal <- tm_shape(hunan)+
  tm_fill('GDPPC',
          n=5,
          style='equal') +
  tm_borders(alpha=0.5) +
  tm_layout(main.title = 'Equal interval classification')

quantile <- tm_shape(hunan)+
  tm_fill('GDPPC',
          n=5,
          style='quantile') +
  tm_borders(alpha=0.5) +
  tm_layout(main.title = 'Equal quantile classification')


```

## 4.4 Global Spatial Autocorrelation

### 4.4.1 Computing Contiguity Spatial Weights

### 4.4.2 Row-standardised weights matrix

### 4.4.3 Global Spatial Autocorrelation: Moran's I

### 4.4.4 Maron's I test

#### 4.4.4.1 Computing Monte Carlo Moran's I

#### 4.4.4.2 Visualising Monte Carlo Moran's I

### 4.4.5 Global Spatial Autocorrelation: Geary's

#### 4.4.5.1 Geary's C test

#### 4.4.5.2 Computing Monte Carlo Geary's C

#### 4.4.5.3 Visualising the Monte Carlo Geary's C

## 4.5 Spatial Correlogram

### 4.5.1 Compute Moran's I correlogram

### 4.5.2 Compute Geary's C correlogram and plot

## 4.6 Cluster and Outlier Analysis

### 4.6.1 Computing local Moran's I

#### 4.6.1.1 Mapping the local Moran's I

#### 4.6.1.2 Mapping local Moran's I values

#### 4.6.1.3 Mapping local Moran's I p-values

#### 4.6.1.4 Mapping both local Moran's I values and p-values

## 4.7 Creating a LISA Cluster Map

### 4.7.1 Plotting Moran scatterplot

### 4.7.2 Plotting Moran scatterplot with standardised variable

### 4.7.3 Preparing LISA map classes

### 4.7.4 Plotting LISA map

## 4.8 Hot Spot and Cold Spot Area Analysis

### 4.8.1 Getis and Ord's G-Statistics

### 4.8.2 Deriving distance-based weight matrix

#### 4.8.2.1 Deriving the centroid

#### 4.8.2.2 Determine the cut-off distance

#### 4.8.2.3 Computing fixed distance weight matrix

### 4.8.3 Computing adaptive distance weight matrix

## 4.9 Computing Gi statistics

### 4.9.1 Gi statistics using fixed distance

### 4.9.2 Mapping Gi values with fixed distance weights

### 4.9.3 Gi statistics using adaptive distance

### 4.9.4 Mapping Gi values with adaptive distance weights
